Implementation of poloidal density variation in collision operator

### 2017-10-05 doc/PoloidalVariationInCollisionOperator_code.tex
updated documentation

\title{Implementation of poloidal density variation in collision operator}

@@

The aim with this document is to explain the modifications in \texttt{SFINCS} that have been done, in order to include poloidal density variation in the Fokker-Planck collision operator.

As is explained in ref.~\cite{ref:PolVarColl_stefan} poloidal density variation can be included by modifying the lowest order distribution function 
\eq{
f_{Ms}(\psi) \rightarrow f_{Ms}(\psi)e^{-Z_se\Phi_1(\theta,\zeta)/T_s},
}
where $f_M(\psi)$ is the flux-function Maxwellian, $\Phi_1$ are first order variation of the electrostatic potential, and $Z_s,T_s$ are the charge and temperature of species $s$ respectively. To implement this change one may simply modify the species density accordingly
\eq{
n_s \rightarrow n_se^{-Ze\Phi_1(\theta,\zeta)/T},
}
in the original equations for the various terms in the collision operator, described in ref.~\cite{ref:Coll1}.

The linearised collision operator takes the form
\eqre{
C_{ab}^{L:f0} =  C_{ab}\left\{f_{aM},f_{bM}\right\}e^{-\left(Z_a/T_a + Z_b/T_b\right)e\Phi_1(\theta,\zeta)} + C_{ab}\left\{f_{a1},f_{bM}\right\}e^{-Z_b/T_be\Phi_1(\theta,\zeta)} + \\ + C_{ab}\left\{f_{aM},f_{b1}\right\}e^{-Z_a/T_ae\Phi_1(\theta,\zeta)},
}
which describes collisions between species $a$ and $b$. In normalised \texttt{SFINCS} units this becomes
\eqre{
\label{eq:CollMod}
\hat C_{ab}^{L:f0} = \hat C_{ab}\left\{f_{aM},f_{bM}\right\}e^{-\left(\hat Z_a/\hat T_a + \hat Z_b/\hat T_b\right) \alpha\hat\Phi_1(\theta,\zeta)} + \hat C_{ab}\left\{f_{a1},f_{bM}\right\}e^{-\hat Z_b/\hat T_b \alpha\hat\Phi_1(\theta,\zeta)} + \\ + \hat C_{ab}\left\{f_{aM},f_{b1}\right\}e^{-\hat Z_a/\hat T_a\alpha\hat\Phi_1(\theta,\zeta)}.
}

\noindent
The first term on the RHS of Eq.~\eqref{eq:CollMod} is the \textit{temperature equilibration term}. Since this term does not include $f_{1s}$ it has to be treated differently compared to the other two cases.

@@

\eqre{
\label{eq:Jac1}
\hat C_{ab}^{L:f0} = -\left(\hat Z_a/\hat T_a + \hat Z_b/\hat T_b\right) \alpha \hat C_{ab}\left\{f_{aM},f_{bM}\right\}e^{-\left(\hat Z_a/\hat T_a + \hat Z_b/\hat T_b\right) \alpha\hat\Phi_1(\theta,\zeta)}  \\ -\hat Z_b/\hat T_b \alpha\ \hat C_{ab}\left\{f_{a1},f_{bM}\right\}e^{-\hat Z_b/\hat T_b \alpha\hat\Phi_1(\theta,\zeta)}  - \hat Z_a/\hat T_a\alpha\hat\Phi_1(\theta,\zeta) \hat C_{ab}\left\{f_{aM},f_{b1}\right\}e^{-\hat Z_a/\hat T_a\alpha\hat\Phi_1(\theta,\zeta)}.
}

\noindent
Apart from the different prefactor, the term in Eq.~\eqref{eq:Jac1} differs from term in $\partial \hat C_{ab}^{L:f0}/\partial \hat f_{1s}$ by acting on the first order distribution function $f_{1s}$. For example, in the first case, terms such as $\partial f_{1s}/\partial x_s$ become 
\eq{
\frac{1}{\partial f_{1s}}\frac{\partial f_{1s}}{\partial x_s} = \frac{\partial}{\partial x_s},
}
whereas in the second case, when take the derivative with respect to $\Phi_1$, we have to actually calculate the derivative $\partial f_{1s}/\partial x_s$.

@@

\section*{Implementation in the Code}
To implement Eq.~\eqref{eq:CollMod} we need to modify the evaluation of the residual and the jacobian. The residual is evaluated in \texttt{evaluateResidual.F90} and the jacobian in \texttt{evaluateJacobian.F90}. Both these two routines call the functions in \texttt{PopulateMatrix.F90}, where the actual assembly of the residual and jacobian matrices are done. Therefore, we only need to modify \texttt{PopulateMatrix.F90}.

\subsection*{Residual}
\label{sec:Res}
In the residual we have to include the extra pre-factor appearing in Eq.~\eqref{eq:CollMod}. This is done by modifying the last block in the calculation of the collision operator, just before the terms are saved into the main matrix. We introduce the pre-factor \texttt{preFactor} which is normally equal to one, but $e^{-\hat Z_a/\hat T_a \alpha\hat\Phi_1(\theta,\zeta)}$ when poloidal density variation is included (by setting the flag \texttt{poloidalVariationInCollisionOperator = .true.}).

\begin{lstlisting}
  preFactor = 1.0 ! Initiate the preFactor used to multiply CHat before saving into the Main matrix
  if (poloidalVariationInCollisionOperator) then 
     preFactor = exp(-Zs(iSpeciesA)*alpha*Phi1Hat(itheta,izeta)/Thats(iSpeciesA)) 
\end{lstlisting}
This factor then multiplies the collision operator, just before saving into the main matrix
\begin{lstlisting}
  call MatSetValueSparse(matrix, rowIndex, colIndex, &
       -nu_n*preFactor*CHat(ix_row,ix_col), ADD_VALUES, ierr) ! Multiply CHat with preFactor before saving
\end{lstlisting}
\subsubsection*{Temperature equilibration term}
In the original version of the code, the temperature equilibration term is calculated in \\ \texttt{evaluateResidual.F90} by using \texttt{whichMatrix = 2} when calling \texttt{PopulateMatrix.F90}. The output is then multiplied with the first order distribution function $f_{0s}$. Since the original version of \texttt{SFINCS} already includes $\Phi_1$ in the definition of $f_{0s}$ (in the subroutine \texttt{init\_f0()} inside \texttt{PopulateMatrix.F90}), the same routines as were presented above can be used. In the end, when calling \texttt{PopulateMatrix} we include the pre-factor $e^{-\hat Z_a/\hat T_a \alpha\hat\Phi_1(\theta,\zeta)}$. Then by multiplying with $f_{0b}$ in \texttt{evaluateResidual.F90} we get the desired total pre-factor $e^{-\left(\hat Z_a/\hat T_a + \hat Z_b/\hat T_b\right) \alpha\hat\Phi_1(\theta,\zeta)}$.

\subsection*{Jacobian}
The $\partial C/\partial f_{1s}$ contribution in the jacobian is at this point already taking the density variations into account if we implement the modifications explained in Section~\ref{sec:Res}. However, we need to add the calculation of $\partial C/\partial \Phi_1$.

\subsubsection*{The $\partial C/\partial \Phi_1$ term}
\label{sec:Jac1}
To implement the derivative with respect to $\Phi_1$ we add another block right after the calculation of the residual terms. This block should be evaluated whenever the jacobian (\texttt{whichmatrix =1}) or the preconditioner (\texttt{whichmatrix =0}) is calculated, with \texttt{poloidalVariationInCollisionOperator = .true.}

\begin{lstlisting}
 if (poloidalVariationInCollisionOperator .and. (whichMatrix == 1 .or. whichMatrix == 0)) then
\end{lstlisting}

We begin with treating the other two terms appearing in Eq.~\eqref{eq:Jac1} and will then look at the temperature equilibration term. For the second two terms in Eq.~\eqref{eq:Jac1} we first have to implement the correct pre-factor.

\begin{lstlisting}
 preFactor = -Zs(iSpeciesA)*alpha/Thats(iSpeciesA)*exp(-Zs(iSpeciesA)&
 *alpha*Phi1Hat(itheta,izeta)/Thats(iSpeciesA))\end{lstlisting}

\noindent
The second change is including the first order distribution function (and calculating derivatives thereof). For this purpose we first have to initiate this distribution function, by reading the appropriate terms in the current state vector

\begin{lstlisting}
 index = getIndex(iSpeciesB,ix,L+1,itheta,izeta,BLOCK_F) ! f1b from statevector
 f1b(ix) = stateArray(index + 1)
\end{lstlisting}
\noindent
What is left is to multiply this vector with the collision operator matrix \texttt{CHat} containing terms such as $\partial/\partial x_s$.

 \begin{lstlisting}
 CHatTimesf = matmul(CHat,f1b)
 \end{lstlisting}
\noindent
The output is then saved into the main matrix in the $\partial \Hat C/\partial \hat \Phi_1$ block,

 \begin{lstlisting}
 call MatSetValue(matrix, rowIndex, colIndex, & 
 -nu_n*preFactor*CHatTimesf(ix_col), ADD_VALUES, ierr) 
\end{lstlisting}
\noindent
using the approptiate row and column indices.

 \begin{lstlisting}
rowIndex=getIndex(iSpeciesA,ix_row,L+1,itheta,izeta,BLOCK_F)
do ix_col = max(ixMinCol,min_x_for_L(L)),Nx
   ! Get column index for the d/dPhi1 terms
   colIndex=getIndex(1,1,1,itheta,izeta,BLOCK_QN)
\end{lstlisting}

\subsubsection*{Temperature equilibration}
\label{sec:Jac2}
If the temperature equilibration terms should be included, we have to add additional terms into the main matrix, when calculating  $\partial \Hat C/\partial \hat \Phi_1$. To begin with we need to include the different pre-factor (since in this case $f_0$ from \texttt{init\_f0()} is not used).

\begin{lstlisting}
if (includeTemperatureEquilibrationTerm) then
   ! Get the correct preFactor
   preFactor = (-Zs(iSpeciesA)*alpha/Thats(iSpeciesA) -Zs(iSpeciesB)*alpha/  & 
    Thats(iSpeciesB)) 
   *exp(-Zs(iSpeciesA)*alpha*Phi1Hat(itheta,izeta)/Thats(iSpeciesA)-Zs(iSpeciesB) & 
   *alpha*Phi1Hat(itheta,izeta)/Thats(iSpeciesB))
\end{lstlisting}
\noindent
The second change originates from the fact that the collision operator in this temperature equilibration case, should multiply the Maxwellian for species $b$ and not $f_{1b}$ (as was previously the case in Section~\ref{sec:Jac2}).  First, we have to initiate this distribution function
 \begin{lstlisting}
 if (includeTemperatureEquilibrationTerm) then
    fM(ix) = expxb2(ix)*nhats(iSpeciesB)*mhats(iSpeciesB)*sqrt(mhats(iSpeciesB)) &
     /(pi*sqrtpi*Thats(iSpeciesB)*sqrt(Thats(iSpeciesB)))
     end if   
\end{lstlisting}
\noindent
and then use it to multiply to the total collision operator.
 \begin{lstlisting}
  CHatTimesf = matmul(CHat,fM)
 \end{lstlisting}
\noindent
Then we add this term into the main matrix, using the the same procedure as was explained in Section~\ref{sec:Jac1}

@@

\bibitem{ref:PolVarColl_stefan} S.~Buller, {\em Poloidal variation in collision operator} (2017).

\bibitem{ref:Coll1} M.~Landreman, {\em Technical Documentation for version 3 of SFINCS} (2014).

%\bibitem{simakov} A.~N.~Simakov, P.~Helander,
%  {\em Phys. Plasmas} {\bf 16}, 042503 (2009).

%\bibitem{nonAxis} P.~Helander, Theory of plasma confinement in non-axisymmetric magnetic fields (2013).

%\bibitem{MH} L.~Råde and B.~Westergren, Mathematics Handbook for Science and Engineering, $5^{\mathrm{th}}$ edition, 2004. %\vspace{-5mm}

%\bibitem{Abra} M.~Abramowitz and I.~A.~Stegun, Handbook of Mathematical Functions, $10^{\mathrm{th}}$ printing, 1972.

\end{thebibliography}

>>>>

\title{Implementation of poloidal density variation in collision operator}
\author{Aylwin Iantchenko}

@@

The aim of this document is to explain the modifications in \texttt{SFINCS} that have been done, in order to include poloidal density variation in the collision operator (in both full Fokker-Planck and pure pitch angle scattering operator options).

As is explained in Ref.~\cite{ref:PolVarColl_stefan} poloidal density variation can be included by modifying the lowest order distribution function 
\eq{
f_{Ms}(\psi) \rightarrow f_{0s} = f_{Ms}(\psi)e^{-Z_se\Phi_1(\theta,\zeta)/T_s},
}
where $f_M(\psi)$ is the flux-function Maxwellian, $\Phi_1$ are first order variation of the electrostatic potential, and $Z_s,T_s$ are the charge and temperature of species $s$ respectively. To implement this change one may simply modify the species density accordingly
\eqre{
\label{eq:densMod}
n_s \rightarrow n_se^{-Ze\Phi_1(\theta,\zeta)/T},
}
in the original equations for the various terms in the collision operator, which are described in Ref.~\cite{ref:Coll1}. The linearised collision operator takes the new form
\eqre{
C_{ab}^{L:f0} =  C_{ab}\left\{f_{aM},f_{bM}\right\}e^{-\left(Z_a/T_a + Z_b/T_b\right)e\Phi_1(\theta,\zeta)} + C_{ab}\left\{f_{a1},f_{bM}\right\}e^{-Z_b/T_be\Phi_1(\theta,\zeta)} + \\ + C_{ab}\left\{f_{aM},f_{b1}\right\}e^{-Z_a/T_ae\Phi_1(\theta,\zeta)},
}
which describes collisions between species $a$ and $b$. In normalised \texttt{SFINCS} units this becomes
\eqre{
\label{eq:CollMod}
\hat C_{ab}^{L:f0} = \hat C_{ab}\left\{f_{aM},f_{bM}\right\}e^{-\left(\hat Z_a/\hat T_a + \hat Z_b/\hat T_b\right) \alpha\hat\Phi_1(\theta,\zeta)} + \hat C_{ab}\left\{f_{a1},f_{bM}\right\}e^{-\hat Z_b/\hat T_b \alpha\hat\Phi_1(\theta,\zeta)} + \\ + \hat C_{ab}\left\{f_{aM},f_{b1}\right\}e^{-\hat Z_a/\hat T_a\alpha\hat\Phi_1(\theta,\zeta)}.
}

\noindent
The first term on the RHS of Eq.~\eqref{eq:CollMod} is the \textit{temperature equilibration term}. Since this term does not include $f_{1s}$ it has to be treated differently compared to the other two cases, as will be seen shortly. In the code, the calculation of the residual has to be modified to include the factors specified in Eq.~\eqref{eq:CollMod}.

@@

\eqre{
\label{eq:Jac1}
\frac{\partial \hat C_{ab}^{L:f0}}{\partial \hat \Phi_1} = -\left(\hat Z_a/\hat T_a + \hat Z_b/\hat T_b\right) \alpha \hat C_{ab}\left\{f_{aM},f_{bM}\right\}e^{-\left(\hat Z_a/\hat T_a + \hat Z_b/\hat T_b\right) \alpha\hat\Phi_1(\theta,\zeta)}  \\ -\hat Z_b/\hat T_b \alpha\ \hat C_{ab}\left\{f_{a1},f_{bM}\right\}e^{-\hat Z_b/\hat T_b \alpha\hat\Phi_1(\theta,\zeta)}  - \hat Z_a/\hat T_a\alpha\hat\Phi_1(\theta,\zeta) \hat C_{ab}\left\{f_{aM},f_{b1}\right\}e^{-\hat Z_a/\hat T_a\alpha\hat\Phi_1(\theta,\zeta)}.
}

\noindent
Apart from the different pre-factor, the term in Eq.~\eqref{eq:Jac1} differs from term in $\partial \hat C_{ab}^{L:f0}/\partial \hat f_{1s}$ by acting on the first order distribution function $f_{1s}$. For example, in the first case, terms such as $\partial f_{1s}/\partial x_s$ become 
\eq{
\frac{1}{\partial f_{1s}}\frac{\partial f_{1s}}{\partial x_s} = \frac{\partial}{\partial x_s},
}
whereas in the second case, when we take the derivative with respect to $\Phi_1$, we have to include the derivative $\partial f_{1s}/\partial x_s$ (and $\partial f_{0s}/\partial x_s$ for temperature equilibration), as well.

@@

\section*{Implementation in the Code}
To implement Eq.~\eqref{eq:CollMod} and Eq.~\eqref{eq:Jac1} we need to modify the evaluation of the residual and the Jacobian. The residual is evaluated in \texttt{evaluateResidual.F90} and the Jacobian in \texttt{evaluateJacobian.F90}. Both these two routines call the functions in \texttt{PopulateMatrix.F90}, where the actual assembly of the residual and Jacobian matrices are done. As a consequence of this structure, we only need to modify \texttt{PopulateMatrix.F90} in order to include poloidal density variation in the collision operator.

\subsection*{Residual}
\label{sec:Res}
In the residual we have to include the extra pre-factor appearing in Eq.~\eqref{eq:CollMod}. Since this factor is species specific, we have to keep track of which species density is appearing in the equations. The pitch-angle scattering and the $\hat C^E_{ab}\left \{f_{a1}\right \}$ parts of the collision operator (see Ref. ~\cite{ref:Coll1}) both uses the density $\texttt{nHats(iSpeciesB)}$ while the other contributions to the total collision operator \texttt{CHat} uses \texttt{nHats(iSpeciesA)} instead.

Rather than creating many if-statements in the current implementation of the collision operator, for sake of clarity we separate the two implementations with one if-statement directly before the various contributions to the collision operator are calculated. If \texttt{poloidalVariationInCollisionOperator = .false.} the first block is executed, which contains the original code.

\begin{lstlisting}
if (.not. poloidalVariationInCollisionOperator) then   !! Added by AI (2017-09).
\end{lstlisting}

\noindent Otherwise, if \texttt{poloidalVariationInCollisionOperator = .true.}, \texttt{includePhi1 = .true.} and \texttt{includePhi1InKineticEquation = .true.} we proceed to the new block where poloidal density variation is included.

\begin{lstlisting}
else if (includePhi1 .and. includePhi1InKineticEquation) then! Do with poloidal density variation
\end{lstlisting}

\noindent
Since we now will have a \texttt{Phi1Hat} dependence in the various terms of the collision operator we will have to iterate over \texttt{itheta} and \texttt{izeta} when calculating the various contributions. Before calculating each term, we define a pre-factor (\texttt{PreFactor}) as is specified in Eq.~\eqref{eq:CollMod}, taking the correct species index into account.

\begin{lstlisting}
 do itheta=ithetaMin,ithetaMax
    do izeta=izetaMin,izetaMax
       ! Generate preFactor for nHats(iSpeciesA) terms
       preFactor =  exp(-Zs(iSpeciesA)*alpha*Phi1Hat(itheta,izeta)/Thats(iSpeciesA))
\end{lstlisting}

\noindent
Because of the \texttt{itheta}, \texttt{izeta} dependence we have redefined \texttt{CECD} and \texttt{nuDHat} to their equivalents \texttt{CECDpol} and \texttt{nuDHatpol} with space for their \texttt{itheta} and \texttt{izeta} components.

The rest of this block is the same as in the original code, except that we do the second iteration over \texttt{itheta}, \texttt{izeta} slightly earlier, and include the appropriate \texttt{PreFactor} in front of additional contributions to the total collision operator \texttt{CHat}.

\subsubsection*{Temperature equilibration term}
In the original version of the code, the temperature equilibration term is calculated in \\ \texttt{evaluateResidual.F90} by using \texttt{whichMatrix = 2} when calling \texttt{PopulateMatrix.F90}. The output is then multiplied with the first order distribution function $f_{0s}$. Since the original version of \texttt{SFINCS} already includes $\Phi_1$ in the definition of $f_{0s}$ (in the subroutine \texttt{init\_f0()} inside \texttt{PopulateMatrix.F90}), the same routines as were presented above can be used also if \texttt{includeTemperatureEquilibrationTerm = .true.}.

\subsection*{Jacobian}
 If the modifications explained in the previous section are implemented, the $\partial C/\partial f_{1s}$ contribution to the Jacobian is at this point already taking the density variations into account. We need to add the calculation of $\partial C/\partial \Phi_1$.

\subsubsection*{The $\partial C/\partial \Phi_1$ term}
\label{sec:Jac1}
We begin with treating the other two terms appearing in Eq.~\eqref{eq:Jac1} and will then look at the temperature equilibration term.

To implement the derivative with respect to $\Phi_1$ we need to redo the calculation of the various contributions to the collision operator, but including a new pre-factor, \texttt{PreFactorJ} as specified in Eq.~\eqref{eq:Jac1}. This should only be done when the Jacobian (\texttt{whichmatrix =1}) or the preconditioner (\texttt{whichmatrix =0}) is calculated.

\begin{lstlisting}
 if (whichMatrix == 1 .or. whichMatrix == 0) then
    ! Generate preFactorJ for nHats(iSpeciesB) terms              
    preFactorJ =  (-Zs(iSpeciesA)*alpha/Thats(iSpeciesA)) &
    *exp(-Zs(iSpeciesA)*alpha*Phi1Hat(itheta,izeta)/Thats(iSpeciesA))
\end{lstlisting}

\noindent
The result is saved in the new matrices \texttt{nuDHatpolJ}, \texttt{CECDpolJ} etc. Same calculations are then repeated for these terms (except for the different pre-factor), and the result is saved in \texttt{CHatJ}.

In contrary to the residual (and the $\partial C/\partial f_{1s}$ terms ), when calculating $\partial C/\partial \Phi_1$ we have to include the first order distribution function before saving the result into the main matrix. For this purpose we first have to initiate this distribution function, by reading the appropriate terms in the current state vector.

\begin{lstlisting}
 do ix= max(ixMinCol,min_x_for_L(L)),Nx
    ! Generate f1b from state vector
    index = getIndex(iSpeciesB,ix,L+1,itheta,izeta,BLOCK_F) 
    f1b(ix) = stateArray(index + 1)
\end{lstlisting}

\noindent
We multiply the terms in the collision operator with this distribution function.

 \begin{lstlisting}
  CHatTimesf = matmul(CHatJ,f1b)
 \end{lstlisting}
\noindent
The result is then saved into the $\partial \Hat C/\partial \hat \Phi_1$ block of the main matrix,

 \begin{lstlisting}
do ix_row=max(ixMin,min_x_for_L(L)),Nx
   rowIndex=getIndex(iSpeciesA,ix_row,L+1,itheta,izeta,BLOCK_F)
   ! Get column index for the d/dPhi1 terms
   colIndex=getIndex(1,1,1,itheta,izeta,BLOCK_QN)
   ! Save into the main matrix, note that here we only use ix_row since CHatTimesf is now a vector
   call MatSetValue(matrix, rowIndex, colIndex, & 
   -nu_n*CHatTimesf(ix_row), ADD_VALUES, ierr) 
   ! need to use MatSetValue, otherwise petsc gives error
end do ! ix_row
\end{lstlisting}
\noindent

\subsubsection*{Temperature equilibration}
\label{sec:Jac2}
If the temperature equilibration terms should be included, we have to add additional terms into the main matrix, when calculating  $\partial \Hat C/\partial \hat \Phi_1$. These terms are different from the other contributions since now we need to multiply the output with $f_{0s} = f_{Ms}(\psi)e^{-Z_se\Phi_1(\theta,\zeta)/T_s}$ (instead of the first order distribution function $f_{1s}$). Because of this, we need to generate the Maxwellian distribution function for \texttt{iSpeciesB}.

\begin{lstlisting}
if (includeTemperatureEquilibrationTerm .and. L==0) then
   fM(ix) = sqrt(mhats(iSpeciesB)/Thats(iSpeciesB))*mhats(iSpeciesB)/Thats(iSpeciesB)&
   * nhats(iSpeciesB)/(pi*sqrtpi)*expxb2(ix)
end if   
\end{lstlisting}

\noindent
Because of the \texttt{Phi1Hat} dependence in the zeroth order distribution function we get a different contribution to the Jacobian, as specified in the first term of Eq.~\eqref{eq:Jac1}. In the code, we need to use both the residual term \texttt{CHat} and the Jacobian terms \texttt{CHatJ} to obtain the correct result.

\begin{lstlisting}
  CHatTimesf = matmul(CHatJ + CHat*(-Zs(iSpeciesB)*alpha/Thats(iSpeciesB))&
   *exp(-Zs(iSpeciesB)*alpha*Phi1Hat(itheta,izeta)/Thats(iSpeciesB)),fM)
\end{lstlisting}

\noindent
What is left is to add this term into the main matrix, using the the same procedure as was explained in the previous section.

\subsection*{Pure pitch angle scattering operator}
The implementation of poloidal density variation in the case the pure pitch angle scattering collision operator is used (corresponding to the case \texttt{collisionoperator = 1}), is slightly easier because of the existing loop over \texttt{itheta}, \texttt{izeta} and since we are working with a scalar \texttt{CHat\_element}. Also, since $L > 0$ always for this term, no temperature equilibration has to be included. The only block we need to change is when we save the collision operator into the main matrix.

We introduce the factor \texttt{preFactor} which is in normal cases equal to one, but when \texttt{poloidalVariationInCollisionOperator = .true.} it is equal to the exponential pre-factor containing $\Phi_1$ in Eq.~\eqref{eq:CollMod}.

Before saving into the main matrix, we multiply the collision operator with this factor.

 \begin{lstlisting}
 call MatSetValue(matrix, index, colIndex, & 
   -nu_n*CHat_element*preFactor, ADD_VALUES, ierr)
  \end{lstlisting}

\subsection*{Jacobian, the $\partial C/\partial \Phi_1$ term}
Just as before, the   $\partial C/\partial f_1$ terms are the same as in the residual. What is different is the $\partial C/\partial \Phi_1$  term. The procedure to calculate this term is however very similar, to the case with the residual. The only difference is the different pre-factor which now also is multiplied with the distribution function from the state vector.

 \begin{lstlisting}
  preFactor = -Zs(iSpeciesA)*alpha/Thats(iSpeciesA)*exp(-Zs(iSpeciesA) &
            *alpha*Phi1Hat(itheta,izeta)/Thats(iSpeciesA))*stateArray(index + 1)
  \end{lstlisting}

\section*{Current state of the project}
Presently, the modifications discussed in the previous section have been implemented into the ``poloidalVariationInCollisionOperator'' branch of \texttt{SFINCS}. What remains is to benchmark the output with theoretical predictions, using for example the model described in Ref.~\cite{ref:Fulop}.

Using the new implementation, for ``reasonable'' input values of impurity charge and main ion temperature gradient, no particular poloidal asymmetries have been observed, and the result is fairly similar if the original code is used instead. Attempts to benchmark with the model described in Ref.~\cite{ref:Fulop} suggests that much larger ion temperature gradients and impurity charge has to be used, in order to observe poloidal asymmetries in the perturbed density. The problem is, at this high gradients and impurity charge, it becomes difficult to converge. In general, it seems that a slightly higher \texttt{Ntheta} has to be used, when including poloidal variations in the collision operator.

@@

\bibitem{ref:PolVarColl_stefan} S.~Buller, \href{https://github.com/landreman/sfincs/blob/poloidalVariationInCollisionOperator/doc/poloidal_variation_in_collision_operator_2017-09-08.pdf}{ \em Poloidal variation in collision operator} (2017).

\bibitem{ref:Coll1} M.~Landreman, \href{https://github.com/landreman/sfincs/blob/poloidalVariationInCollisionOperator/doc/20150507-01\%20Technical\%20documentation\%20for\%20version\%203\%20of\%20SFINCS.pdf}{\em Technical Documentation for version 3 of SFINCS} (2014).

\bibitem{ref:Fulop} T.~F\"{u}l\"{o}p, P.~Helander \href{http://aip.scitation.org/doi/abs/10.1063/1.873593}{\em Phys. Plasmas}{ \bf 6}, 3066 (1999) 
\end{thebibliography}
